Supplement to Multi-scale trend analysis of water quality using error propagation of generalized additive models

### 2021-05-05 manu_draft.Rmd
5 and 10 year trend comparisons added to supplement

data(locs)
data(modprf)
data(seastrnd)
data(seastrnd2)
data(cmptrnd)
data(modstr)
data(modslog_chl18)
data(modslog_chl24)
data(modslog_chl27)
data(modslog_chl32)
data(modslog_chl34)

@@

Applying the same comparison to all stations showed that different trend analysis methods provided conflicting information on the magnitude and significance of the seasonal chl-a changes in each decade (Figure \@ref(fig:trndcmp)).  In many cases, the slope estimates were similar in magnitude, with some exceptions at the more southern stations where the OLS estimates suggested a larger trend than the meta-analysis methods. More importantly, differences in the magnitude of the confidence intervals between the OLS models applied to the GAM averages and the meta-analyses were also observed, reflecting the ability of the latter to more accurately assess significance of trends by accounting for uncertainty in the average estimates.

@@

Incorrect conclusions on trends can have dramatic consequences for regulated parties under existing water quality compliance frameworks [@Smith01]. Our examples in Figures \@ref(fig:trndcmpex) and \@ref(fig:trndcmp) demonstrate these risks if propagation of uncertainty from raw observations across methods is unaccounted for in trend assessment.  Our assessment of trends using OLS regression applied to seasonal averages from the raw observations is effectively like averaging results within a year and applying a simple Kendall test. In many cases the results may be similar, but loss of information with averaging can lead to increased Type I or II error rates depending on characteristics of the raw data and the method used for their evaluation [@Shabman03]. Our examples demonstrated the potential for incorrect conclusions at specific monitoring locations and at much larger spatial scales across all stations.

@@

```{r, results = 'hide', message = F}
toplo <- cmptrnd
p <- ggplot(toplo, aes(y = station, x = yrcoef, linetype = pval, shape = modtyp, colour = modtyp)) + 
  facet_grid(seas ~ yrs, scales = 'free_x') + 
  geom_point(size = 2, position = position_dodge(width = 0.6), alpha = 0.8) + 
  geom_errorbarh(aes(xmin = yrcoeflwr, xmax = yrcoefupr), position = position_dodge(width = 0.6), height = 0) +
  scale_linetype_manual('Trend significance', values = c('dashed', 'solid')) + 
  scale_colour_discrete_qualitative(name = 'Trend model', palette = 'Dark3') +
  scale_shape(name = 'Trend model') + 
  geom_vline(xintercept = 0, linetype = 'dashed') + 
  # scale_x_continuous(limits = c(-1.14, 1.14)) +
  guides(
    color = guide_legend(override.aes = list(linetype = 0))
  ) +
  theme_minimal() + 
  theme(
    strip.background = element_blank(),
    panel.background = element_rect(fill = 'gray98', colour = NA), 
    panel.border = element_blank(), 
    legend.position = 'top'
  ) +
  labs(
    y = 'Station (north to south)', 
    x = expression(paste(log[10], " chl-a change (", italic(mu), "g ", L^-1, yr^-1, ")"))
  )

@@

# save as png
png('figs/trndcmp.png', height = 6, width = 7.5, family = 'serif', units = 'in', res = 400)
p
dev.off()
```
```{r trndcmp, fig.cap = 'Trend estimate comparisons for three models applied to seasonal averages of chl-a in different annual periods at each station. The "OLS raw" trend model is based on an ordinary least squares (OLS) regression fit to the seasonal averages of chl-a from the raw data, the "OLS GAM" trend model is based on an OLS regression fit to the seasonal averages of chl-a from the GAM model (without error propagation), and the "Meta-analysis GAM" trend model is based on a meta-analysis regression fit to the seasonal averages of chl-a from the GAM model.  Values for each model are the log$_{10}$-slope estimates (+/- 95\\% confidence interval) as annual change per year within each season, with line style denoting trend significance.'}
knitr::include_graphics('figs/trndcmp.png')
```

>>>>

data(locs)
data(modprf)
data(seastrnd)
data(seastrnd2)
data(cmptrnd10)
data(modstr)
data(modslog_chl18)
data(modslog_chl24)
data(modslog_chl27)
data(modslog_chl32)
data(modslog_chl34)

@@

Applying the same comparison to all stations showed that different trend analysis methods provided conflicting information on the magnitude and significance of the seasonal chl-a changes in each decade (Figure \@ref(fig:trndcmp10), see supplement for the same analysis with five and fifteen year time periods).  In many cases, the slope estimates were similar in magnitude, with some exceptions at the more southern stations where the OLS estimates suggested a larger trend than the meta-analysis methods. More importantly, differences in the magnitude of the confidence intervals between the OLS models applied to the GAM averages and the meta-analyses were also observed, reflecting the ability of the latter to more accurately assess significance of trends by accounting for uncertainty in the average estimates.

@@

Incorrect conclusions on trends can have dramatic consequences for regulated parties under existing water quality compliance frameworks [@Smith01]. Our examples in Figures \@ref(fig:trndcmpex) and \@ref(fig:trndcmp10) demonstrate these risks if propagation of uncertainty from raw observations across methods is unaccounted for in trend assessment.  Our assessment of trends using OLS regression applied to seasonal averages from the raw observations is effectively like averaging results within a year and applying a simple Kendall test. In many cases the results may be similar, but loss of information with averaging can lead to increased Type I or II error rates depending on characteristics of the raw data and the method used for their evaluation [@Shabman03]. Our examples demonstrated the potential for incorrect conclusions at specific monitoring locations and at much larger spatial scales across all stations.

@@

```{r, results = 'hide', message = F}
toplo <- cmptrnd10
p <- ggplot(toplo, aes(y = station, x = yrcoef, linetype = pval, shape = modtyp, colour = modtyp)) + 
  facet_grid(seas ~ yrs, scales = 'free_x') + 
  geom_point(size = 2, position = position_dodge(width = 0.6), alpha = 0.8) + 
  geom_errorbarh(aes(xmin = yrcoeflwr, xmax = yrcoefupr), position = position_dodge(width = 0.6), height = 0) +
  scale_linetype_manual('Trend significance', values = c('dashed', 'solid')) + 
  scale_colour_discrete_qualitative(name = 'Trend model', palette = 'Dark3') +
  scale_shape(name = 'Trend model') + 
  geom_vline(xintercept = 0, linetype = 'dashed') + 
  # scale_x_continuous(limits = c(-1.14, 1.14)) +
  guides(
    color = guide_legend(override.aes = list(linetype = 0))
  ) +
  theme_minimal() + 
  theme(
    strip.background = element_blank(),
    panel.background = element_rect(fill = 'gray98', colour = NA), 
    panel.border = element_blank(), 
    legend.position = 'top'
  ) +
  labs(
    y = 'Station (north to south)', 
    x = expression(paste(log[10], " chl-a change (", italic(mu), "g ", L^-1, yr^-1, ")"))
  )

@@

# save as png
png('figs/trndcmp10.png', height = 6, width = 7.5, family = 'serif', units = 'in', res = 400)
p
dev.off()
```
```{r trndcmp10, fig.cap = 'Trend estimate comparisons for three models applied to seasonal averages of chl-a in different annual periods at each station. The "OLS raw" trend model is based on an ordinary least squares (OLS) regression fit to the seasonal averages of chl-a from the raw data, the "OLS GAM" trend model is based on an OLS regression fit to the seasonal averages of chl-a from the GAM model (without error propagation), and the "Meta-analysis GAM" trend model is based on a meta-analysis regression fit to the seasonal averages of chl-a from the GAM model.  Values for each model are the log$_{10}$-slope estimates (+/- 95\\% confidence interval) as annual change per year within each season, with line style denoting trend significance.'}
knitr::include_graphics('figs/trndcmp10.png')
```

### 2021-05-05 supplement.Rmd
5 and 10 year trend comparisons added to supplement

data(locs)
data(modprf)
data(seastrnd)
data(seastrnd2)
data(cmptrnd)
data(modstr)
data(modslog_chl18)
data(modslog_chl24)
data(modslog_chl27)
data(modslog_chl32)
data(modslog_chl34)

@@

>>>>

data(locs)
data(modprf)
data(seastrnd)
data(seastrnd2)
data(cmptrnd05)
data(cmptrnd15)
data(modstr)
data(modslog_chl18)
data(modslog_chl24)
data(modslog_chl27)
data(modslog_chl32)
data(modslog_chl34)

@@

```{r, results = 'hide', message = F}
toplo <- cmptrnd05
p <- ggplot(toplo, aes(y = station, x = yrcoef, linetype = pval, shape = modtyp, colour = modtyp)) + 
  facet_grid(seas ~ yrs, scales = 'free_x') + 
  geom_point(size = 2, position = position_dodge(width = 0.6), alpha = 0.8) + 
  geom_errorbarh(aes(xmin = yrcoeflwr, xmax = yrcoefupr), position = position_dodge(width = 0.6), height = 0) +
  scale_linetype_manual('Trend significance', values = c('dashed', 'solid')) + 
  scale_colour_discrete_qualitative(name = 'Trend model', palette = 'Dark3') +
  scale_shape(name = 'Trend model') + 
  geom_vline(xintercept = 0, linetype = 'dashed') + 
  # scale_x_continuous(limits = c(-1.14, 1.14)) +
  guides(
    color = guide_legend(override.aes = list(linetype = 0))
  ) +
  theme_minimal() + 
  theme(
    strip.background = element_blank(),
    panel.background = element_rect(fill = 'gray98', colour = NA), 
    panel.border = element_blank(), 
    legend.position = 'top', 
    axis.text.x = element_text(angle = 45, hjust = 1)
  ) +
  labs(
    y = 'Station (north to south)', 
    x = expression(paste(log[10], " chl-a change (", italic(mu), "g ", L^-1, yr^-1, ")"))
  )

# save as png
png('figs/trndcmp05.png', height = 6, width = 7.5, family = 'serif', units = 'in', res = 400)
p
dev.off()
```
```{r trndcmp05, fig.cap = 'Trend estimate comparisons for three models applied to seasonal averages of chl-a in 5 year periods at each station. The "OLS raw" trend model is based on an ordinary least squares (OLS) regression fit to the seasonal averages of chl-a from the raw data, the "OLS GAM" trend model is based on an OLS regression fit to the seasonal averages of chl-a from the GAM model (without error propagation), and the "Meta-analysis GAM" trend model is based on a meta-analysis regression fit to the seasonal averages of chl-a from the GAM model.  Values for each model are the log$_{10}$-slope estimates (+/- 95\\% confidence interval) as annual change per year within each season, with line style denoting trend significance.'}
knitr::include_graphics('figs/trndcmp05.png')
```

```{r, results = 'hide', message = F}
toplo <- cmptrnd15
p <- ggplot(toplo, aes(y = station, x = yrcoef, linetype = pval, shape = modtyp, colour = modtyp)) + 
  facet_grid(seas ~ yrs, scales = 'free_x') + 
  geom_point(size = 2, position = position_dodge(width = 0.6), alpha = 0.8) + 
  geom_errorbarh(aes(xmin = yrcoeflwr, xmax = yrcoefupr), position = position_dodge(width = 0.6), height = 0) +
  scale_linetype_manual('Trend significance', values = c('dashed', 'solid')) + 
  scale_colour_discrete_qualitative(name = 'Trend model', palette = 'Dark3') +
  scale_shape(name = 'Trend model') + 
  geom_vline(xintercept = 0, linetype = 'dashed') + 
  # scale_x_continuous(limits = c(-1.14, 1.14)) +
  guides(
    color = guide_legend(override.aes = list(linetype = 0))
  ) +
  theme_minimal() + 
  theme(
    strip.background = element_blank(),
    panel.background = element_rect(fill = 'gray98', colour = NA), 
    panel.border = element_blank(), 
    legend.position = 'top'
  ) +
  labs(
    y = 'Station (north to south)', 
    x = expression(paste(log[10], " chl-a change (", italic(mu), "g ", L^-1, yr^-1, ")"))
  )

# save as png
png('figs/trndcmp15.png', height = 6, width = 7.5, family = 'serif', units = 'in', res = 400)
p
dev.off()
```
```{r trndcmp15, fig.cap = 'Trend estimate comparisons for three models applied to seasonal averages of chl-a in 15 year periods at each station. The "OLS raw" trend model is based on an ordinary least squares (OLS) regression fit to the seasonal averages of chl-a from the raw data, the "OLS GAM" trend model is based on an OLS regression fit to the seasonal averages of chl-a from the GAM model (without error propagation), and the "Meta-analysis GAM" trend model is based on a meta-analysis regression fit to the seasonal averages of chl-a from the GAM model.  Values for each model are the log$_{10}$-slope estimates (+/- 95\\% confidence interval) as annual change per year within each season, with line style denoting trend significance.'}
knitr::include_graphics('figs/trndcmp15.png')
```

# References
